---
title: "Analysis of UCSB Undergraduate Grade Trends"
author: "Shreya Sinha" # <- add your name here
date: "`r Sys.Date()`"
params: 
  showCode: FALSE
execute: 
  warning: FALSE # <- this makes sure your code chunks are evaluated
  message: FALSE
format:
  html:
    fontsize: 1.1rem
fig-cap-location: top
output: pdf_document
---

```{r setup, include = FALSE}
rm(list = ls())
library(tidyverse)
library(janitor)
library(lubridate)
library(gt)
library(grid)
library(data.table)
source("cleaning_grades.R")
```

### Data

UC Santa Barbara students’ grade data from Fall 2009 to Winter 2025 were obtained through Public Records Act requests facilitated by the Daily Nexus. The dataset includes detailed information on course grades across various departments and academic levels, encompassing both graduate and undergraduate courses. It provides a breakdown of letter grades, Pass/No Pass enrollments, and other grade-related metrics.

### Findings

Figure 1 shows the average GPA for undergraduate courses across the College of Engineering, College of Letters & Science, Gervitz School of Education, and the three divisions within Letters & Science, spanning the 2009-2010 to 2024-2025 academic years (excluding Spring and Summer 2025). Over this period, a general upward trend in average GPA is evident across all colleges and divisions, with the most significant increases observed in the Humanities & Fine Arts division of Letters & Science and the Gervitz School of Education. In the 2021-2022 academic year, all divisions and colleges, except for Humanities & Fine Arts and the Gervitz School of Education, experienced a marked decline in average GPA.

Figure 2 complements this analysis by showing the percentage of A's awarded in undergraduate courses across the same colleges and divisions over the same period. This percentage includes A plus and minus grades and mirrors the overall trend in average GPA. The most notable increase in the percentage of A's is in the Mathematical, Life, & Physical Sciences division of Letters & Science, which has approximately doubled since the 2009-2010 academic year.

```{r, echo = FALSE, fig.width=9, fig.height=4}
#| fig-cap: "Figure 1: Average GPA Over Time for UCSB Colleges and Divisions"

college_division_gpa_plot
```

\newpage

```{r, echo = FALSE, fig.width=9, fig.height=4}
#| fig-cap: "Figure 2: Percentage of A's Over Time for UCSB Colleges and Divisions"

college_division_A_plot
```

The percentage of A grades awarded in undergraduate courses in the 2023-2024 academic year for each department is presented on the last page.

### Methods

Each department in the dataset was manually categorized into its respective division and college, as this information was not available in the original dataset. The division and college assignments were based on existing academic classifications at UC Santa Barbara, with departments categorized according to their field of study. This categorization was essential for the subsequent analysis of grading trends across different academic units. To facilitate analysis of undergraduate course grades, courses with course numbers 200 or higher were excluded to remove graduate-level courses.

To analyze grade trends across colleges and academic divisions, data was aggregated in two stages: first at the course level and then at the college and division levels. For courses taught by multiple instructors or offered across multiple quarters within an academic year, the course-level average GPA was calculated as a weighted mean, with weights based on the total number of students enrolled in each course. Additionally, the total enrollment and the total number of A's (including plus and minus grades) for each course were summed across all quarters in the academic year.

Using these course-level metrics, data was then aggregated at the college and division levels for each academic year. For both colleges and divisions, the average GPA was calculated as a weighted mean of the course-level GPAs, with weights derived from yearly course enrollments. The total enrollment for each college and division was determined by summing the enrollments of all courses within the respective grouping. The percentage of A grades awarded was calculated by dividing the total number of A grades by the total enrollment for each college or division. This approach ensured a consistent methodology for examining grading patterns while accounting for variability in course sizes, grading distributions, and enrollment trends across academic groupings. \newpage

```{r echo = FALSE, message=FALSE, warning=FALSE}

percent_ls_full_pretty 

percent_rest
```
